\chapter{Experiments}
\label{ch:experiments}
\label{ch:evaluation}

This chapter evaluates RA-GARK on the sparse review-derived KG introduced in Chapter~\ref{ch:intro}. Section~\ref{sec:eval-dataset} summarizes the dataset and the KG construction pipeline, which we adopt from prior work rather than propose here. Section~\ref{sec:eval-setup} specifies the evaluation protocol and the baseline reproduction settings. Sections~\ref{sec:eval-main} and~\ref{sec:eval-ablation} report the main results and the per-component ablation study. Sections~\ref{sec:eval-sensitivity}--\ref{sec:eval-seed} examine the two axes most likely to concern a reader: hyperparameter sensitivity and seed stability. Section~\ref{sec:eval-case} presents an item-level case study of the rationale module, and Section~\ref{sec:eval-cost} reports computational cost.

\section{Dataset and KG Construction}
\label{sec:eval-dataset}

\subsection{Dataset}
\label{sec:eval-dataset-stats}

Our experiments use a filtered subset of the Amazon Books review corpus, represented as a user--item interaction graph and an item--aspect knowledge graph. Table~\ref{tab:dataset-stats} summarizes the post-filtering statistics. After removing low-engagement users and items and pruning rare aspect tags, the working dataset contains $905$ users, $1{,}399$ items, $22{,}265$ positive interactions, and $3{,}370$ KG edges across $2{,}098$ unique aspect tags. On average, each item carries $2.4$ aspect edges, a sparsity regime in which all four KG-aware baselines we tested are outperformed by pure LightGCN (Chapter~\ref{ch:intro}).

\begin{table}[!htbp]
\centering
\caption{Dataset statistics after preprocessing.}
\label{tab:dataset-stats}
\begin{tabular}{lr}
\toprule
Statistic & Value \\
\midrule
\#Users                            & $905$ \\
\#Items                            & $1{,}399$ \\
\#Interactions                     & $22{,}265$ \\
\#KG edges (after filtering)       & $3{,}370$ \\
\#Unique aspects                   & $2{,}098$ \\
Avg.\ edges per item               & $2.4$ \\
Interaction density                & $1.76\%$ \\
\bottomrule
\end{tabular}
\end{table}
\FloatBarrier

Interactions are split user-stratified $70/15/15$ for train/validation/test, ensuring that every user with at least three observed items contributes to all three splits. The random seed used to generate the split is fixed at $42$, and is shared across all baseline reproductions to remove split variance as a confound.

\subsection{Knowledge Graph Construction}
\label{sec:eval-dataset-kg}

The item--aspect KG used here is constructed from raw product reviews by the pipeline of~\cite{ho2024reviewkg}, and we adopt it without modification. Because this construction is not a contribution of this thesis, we only summarize the steps needed to interpret the resulting graph. For each book, visual features are extracted from the cover image with ResNet and Qwen-VL, and a unified textual summary is generated from the review pool with BART and Mistral. Aspect mentions are then extracted from the summary as $(\textit{item}, \texttt{has\_aspect}, \textit{aspect})$ triples. The raw extraction yields $13{,}905$ candidate edges.

Two filtering steps reduce this to the working KG. First, we remove the top $2\%$ most frequent aspect tags, which empirically collapse to generic praise terms such as ``well-written'' or ``engaging''; such tags carry little item-discriminating information and otherwise dominate the SVD spectrum (Section~\ref{sec:kgsvd}). Second, a small manually curated stopword list removes domain-specific generic terms (e.g.\ ``character\_development'', ``highly\_recommended''). Together, the two passes drop $75.7\%$ of the raw edges, leaving $3{,}370$ specific aspect attachments across $2{,}098$ unique tags.

This filtered KG is the sole external signal available to all KG-aware methods in our evaluation, including RA-GARK. The same filtering is applied to every method so that differences in performance reflect modeling choices rather than differences in KG cleanliness.

\section{Experimental Setup}
\label{sec:eval-setup}

\subsection{Baselines}
\label{sec:eval-baselines}

We compare RA-GARK against five baselines: pure LightGCN~\cite{he2020lightgcn} as the non-KG anchor, and four representative KG-aware recommenders, KGAT~\cite{wang2019kgat}, KGCL~\cite{yang2022kgcl}, MCCLK~\cite{zou2022mcclk}, and KGRec~\cite{yang2023kgrec}. These KG-aware baselines cover the main integration paradigms surveyed in Chapter~\ref{ch:relatedwork}: deep aggregator fusion (KGAT), contrastive cross-view learning (KGCL, MCCLK), and edge-level rationale learning (KGRec).

All baselines are reproduced from the authors' released code on the same processed KG (Section~\ref{sec:eval-dataset-kg}) and the same user-stratified split. We keep each baseline's published hyperparameter defaults rather than retuning per method, because retuning four distinct hyperparameter spaces would introduce comparison artifacts; the original authors are best positioned to know their own model's preferred configuration. The only intervention is to fix the embedding dimension at $d = 128$ across all methods, matching RA-GARK, and to align the optimization budget (epochs and early-stopping patience) with our own. These choices are intentionally conservative: they may understate the baselines' achievable performance on our dataset, but they keep the comparison free of uneven tuning effort.

\subsection{Evaluation Protocol}
\label{sec:eval-protocol}

We evaluate under the full-ranking protocol: for each test user, we score every item in $\mathcal{I}$ except those already observed in the user's training set, rank the resulting candidate list, and compute top-$K$ metrics against the held-out test positives. This is the strict protocol; sampled-evaluation variants (for example, ranking against $100$ random negatives) are not used, because they have been shown to overstate model differences in implicit-feedback settings.

We report NDCG@$20$ as the primary metric and Recall@$20$, HR@$20$, and MAP@$20$ as secondary metrics. Early stopping is performed on validation NDCG@$20$ with patience $10$, and the test metrics are taken from the epoch with the best validation NDCG@$20$. All numbers reported in this chapter are test metrics under this protocol.

\subsection{Hyperparameter Settings}
\label{sec:eval-hparams}

RA-GARK's hyperparameters are summarized in Table~\ref{tab:hparams}. These values are not jointly tuned: the optimizer settings and embedding dimension follow the LightGCN default; the rationale temperature $\tau = 0.5$ and fusion-gate bias $b = +5$ are the design choices justified in Sections~\ref{sec:softmax-softmax} and~\ref{sec:gate-bias}; the number of aspect slots $A = 4$ is fixed by the design rationale of Section~\ref{sec:kgsvd-slots}; and the contrastive weight $\lambda_{\mathrm{CL}} = 0.005$ is the small auxiliary value justified in Section~\ref{sec:cl-lambda}.

\begin{table}[!htbp]
\centering
\caption{RA-GARK hyperparameter settings used in experiments.}
\label{tab:hparams}
\begin{tabular}{ll}
\toprule
Hyperparameter & Value \\
\midrule
Embedding dimension $d$            & $128$ \\
LightGCN layers $K$                & $2$ \\
Aspect slots $A$                   & $4$ \\
Rationale temperature $\tau$       & $0.5$ \\
Fusion-gate bias $b$ (init)        & $+5$ \\
Contrastive weight $\lambda_{\mathrm{CL}}$ & $0.005$ \\
InfoNCE temperature $\tau_{\mathrm{CL}}$   & $0.2$ \\
Optimizer                          & Adam ($\beta_1{=}0.9$, $\beta_2{=}0.999$) \\
Learning rate                      & $10^{-3}$ \\
Batch size                         & $128$ \\
Max epochs                         & $80$ \\
Early-stopping patience            & $10$ \\
Random seed                        & $42$ \\
\bottomrule
\end{tabular}
\end{table}
\FloatBarrier

\subsection{Hardware and Software}
\label{sec:eval-env}

To make the experimental setting reproducible, all experiments are conducted on a single NVIDIA GeForce RTX 3090 GPU with $24$\,GB of memory. The model is implemented in PyTorch~2.6.0 with CUDA~12.6 support and trained in FP32. Per-epoch wall-clock time is reported in Section~\ref{sec:eval-cost}.

\section{Main Results}
\label{sec:eval-main}

Table~\ref{tab:main} reports the main test-set results. RA-GARK attains NDCG@$20 = 0.1242$, improving on the strongest KG-aware baseline (KGRec) by $+13.4\%$ and on pure LightGCN by $+5.3\%$. The improvement is consistent across the secondary ranking metrics (HR@$20$, Recall@$20$, and MAP@$20$); the largest relative gain over KGRec is $+18.6\%$ on MAP@$20$, indicating that the rank quality of correctly retrieved items is also better than the baselines'.

\begin{table}[!htbp]
\centering
\caption{Comparison of recommendation performance under Top-20 evaluation.}
\label{tab:main}
\begin{tabular}{lcccc}
\toprule
Model & NDCG@20 & HR@20 & Recall@20 & MAP@20 \\
\midrule
MCCLK~\cite{zou2022mcclk}   & $0.1067$ & $0.4530$ & $0.1720$ & $0.0497$ \\
KGCL~\cite{yang2022kgcl}    & $0.1073$ & $0.4696$ & $0.1827$ & $0.0479$ \\
KGAT~\cite{wang2019kgat}    & $0.1079$ & $0.4773$ & $0.1807$ & $0.0491$ \\
KGRec~\cite{yang2023kgrec}  & $0.1095$ & $0.4729$ & $0.1834$ & $0.0500$ \\
LightGCN~\cite{he2020lightgcn} (no KG) & $0.1179$ & $0.4917$ & $0.1937$ & $0.0555$ \\
\midrule
\textbf{RA-GARK (ours)}     & $\mathbf{0.1242}$ & $\mathbf{0.4983}$ & $\mathbf{0.2022}$ & $\mathbf{0.0593}$ \\
\midrule
$\Delta$ vs KGRec            & $+13.4\%$ & $+5.4\%$  & $+10.3\%$ & $+18.6\%$ \\
$\Delta$ vs LightGCN         & $+5.3\%$  & $+1.3\%$  & $+4.4\%$  & $+6.8\%$  \\
\bottomrule
\end{tabular}
\end{table}
\FloatBarrier

Two readings of this table are worth separating. The first, $+13.4\%$ over KGRec, shows that, with the underlying KG held fixed, RA-GARK's design choices yield a substantially better recommender than the strongest published KG-aware method on the same data. The second, $+5.3\%$ over pure LightGCN, is the methodologically more interesting one. In this sparse-KG setting, all four KG-aware baselines fall below the no-KG anchor; RA-GARK is the only one of the five KG-aware configurations evaluated here that turns the KG signal from a net liability into a net asset. The architectural mechanism responsible for this flip, namely gating the KG channel behind a bias-initialized fusion gate so that it is engaged only when its contribution reduces loss, is examined in detail in the next section.

\section{Ablation Study}
\label{sec:eval-ablation}
\label{sec:ablation-svd}
\label{sec:ablation-softmax}
\label{sec:ablation-gate}

We measure each of the three headline design choices, KG-SVD initialization (Section~\ref{sec:kgsvd}), softmax rationale masking (Section~\ref{sec:softmax}), and the locally biased fusion gate (Section~\ref{sec:gate}), by individually reverting it and retraining from scratch with all other settings held fixed. We additionally include three diagnostic rows: removing the global view entirely (CL-only dual view), replacing the per-(user, item) gate with a single learnable scalar $\alpha$, and dropping each contrastive channel individually. Pure LightGCN is reproduced as the no-KG floor. All rows use seed $= 42$, and $\Delta$ denotes the relative NDCG@20 change against the full model.

\begin{table}[!htbp]
\centering
\caption{Ablation results for RA-GARK design components.}
\label{tab:ablation}
\begin{tabular}{lcc l}
\toprule
Configuration & NDCG@20 & $\Delta$ & What is removed \\
\midrule
\textbf{RA-GARK (full)}              & $\mathbf{0.1242}$ & $0\%$       & none \\
\midrule
$-$ Softmax $\to$ sigmoid head       & $0.1049$ & $-15.5\%$ & cross-aspect competition (\S\ref{sec:softmax-vs-sigmoid}) \\
$-$ KG-SVD init $\to$ Xavier         & $0.1169$ & $-5.9\%$  & angular geometry at init (\S\ref{sec:kgsvd}) \\
$-$ Fusion-gate bias $b{=}5 \to 0$   & $0.1186$ & $-4.5\%$  & local-biased safe default (\S\ref{sec:gate-bias}) \\
$-$ MLP gate $\to$ scalar $\alpha$   & $0.1183$ & $-4.7\%$  & per-(u,i) gate conditioning (\S\ref{sec:gate-perside}) \\
$-$ User CL ($\mathcal{L}_{\mathrm{uCL}}$)  & $0.1189$ & $-4.3\%$  & user-side cross-view alignment \\
$-$ Aspect CL ($\mathcal{L}_{\mathrm{aCL}}$) & $0.1204$ & $-3.1\%$  & item-side aspect-level alignment \\
$-$ Global view (CL-only dual view)  & $0.1218$ & $-1.9\%$  & whole KG pipeline at score time \\
$-$ Rationale enabled $\to$ uniform mean & $0.1235$ & $-0.6\%$  & user-conditioned aspect selection \\
\midrule
LightGCN only (no-KG floor)          & $0.1179$ & ref.       & no global view; CF only \\
Pre-fix configuration (sigmoid + $b{=}0$) & $0.1050$ & $-15.5\%$ & both broken defaults reverted \\
\bottomrule
\end{tabular}
\end{table}
\FloatBarrier

Table~\ref{tab:ablation} supports three observations.

\paragraph{All four targeted components contribute positively.} Reverting any one of the softmax rationale ($-15.5\%$), KG-SVD initialization ($-5.9\%$), local-biased fusion bias ($-4.5\%$), or the per-(u,i) MLP gate ($-4.7\%$) lowers NDCG@$20$ below the full model. The two largest of these, softmax ($0.1049$) and KG-SVD ($0.1169$), additionally fall below the pure-LightGCN floor of $0.1179$: without either component, RA-GARK is no longer a positive-contribution KG-aware method on this dataset. The fusion-bias and gate-conditioning reverts ($0.1186$ and $0.1183$) stay marginally above the floor but still clearly below the full model, so none of the four design choices is decorative.

\paragraph{The softmax-vs-sigmoid choice is by far the largest single-component effect.} The $-15.5\%$ NDCG@$20$ drop when the cross-aspect softmax is replaced with an element-wise sigmoid (to $0.1049$) is by a wide margin the largest individual ablation in the table, several times the typical $1$--$3\%$ swing produced by ordinary hyperparameter tuning. This is the empirical basis for the methodological observation we develop in Chapter~\ref{ch:conclusion}: in our rationale module, the choice of attention normalization is not a tuning detail but a design choice tied to the rationale's semantic assumption, namely that the $A$ slots compete with one another rather than contribute independently.

\paragraph{The pre-fix configuration recovers the broken baseline behavior.} Reverting both fixes simultaneously, sigmoid rationale together with $b = 0$, gives NDCG@$20 = 0.1050$, which is below \emph{all} four KG-aware baselines. This shows that the architectural skeleton of RA-GARK (two parallel views plus late fusion) is not sufficient by itself: stripped of the headline design choices, the same skeleton reproduces the failure mode exhibited by the baselines, exactly as expected from the design principle in Section~\ref{sec:approach-principle}. The two reverts do not stack additively: the sigmoid revert alone ($0.1049$) is already as damaging as the two combined ($0.1050$), indicating that the fusion-gate bias matters only once a softmax head is present to produce a usable rationale; with a sigmoid head the channel is already degraded regardless of the gate-bias setting.

The contrastive losses contribute smaller but non-trivial amounts: $-4.3\%$ for user-side CL and $-3.1\%$ for aspect-level CL. We interpret these magnitudes as consistent with the auxiliary role assigned to contrastive regularization in Section~\ref{sec:cl-role}: they nudge the geometry, but they do not carry the integration load.

\section{Sensitivity Analysis}
\label{sec:eval-sensitivity}

We probe RA-GARK's sensitivity along the four hyperparameter axes that a reader is most likely to question: the softmax temperature $\tau$, the number of latent aspect slots $A$, the fusion-gate bias $b$, and the contrastive weight $\lambda_{\mathrm{CL}}$. Each axis is swept with all other settings held at the values of Table~\ref{tab:hparams} and seed $= 42$. The best configuration, NDCG@$20 = 0.1242$, anchors the section; duplicate default settings cached under sweep-specific tags can differ slightly at the fourth decimal place, so we treat such duplicates as equivalent.

\subsection{Softmax Temperature $\tau$}
\label{sec:sens-tau}

Table~\ref{tab:sens-tau} reports NDCG@$20$ across $\tau \in \{0.05, 0.1, 0.5\}$. The default $\tau = 0.5$ remains the best setting, with a sweep-specific duplicate run at $0.1240$ and the main winner row at $0.1242$. Sharpening to $\tau = 0.1$ costs about $-1.0\%$ ($0.1229$) and to $\tau = 0.05$ costs about $-1.9\%$ ($0.1219$). The picture is consistent with the analysis in Section~\ref{sec:softmax-softmax}: at $\tau = 0.5$ the slot logits are amplified enough to produce discriminative weights without collapsing onto a single slot; pushing $\tau$ down toward $0.05$ drives the rationale toward an approximately one-hot selection, which on our softly-peaked weights (Section~\ref{sec:eval-case}) discards useful mass on the secondary slots and costs a small amount of NDCG@$20$. In the opposite direction, raising $\tau$ toward $1$ flattens the softmax toward a uniform item-wise mean over slots; we keep $\tau = 0.5$ as the value that balances these two regimes.

\begin{table}[!htbp]
\centering
\caption{Sensitivity to softmax temperature $\tau$.}
\label{tab:sens-tau}
\begin{tabular}{lc}
\toprule
$\tau$ & NDCG@20 \\
\midrule
$\mathbf{0.5}$ (default) & $\mathbf{0.1240}$ \\
$0.1$              & $0.1229$ \\
$0.05$             & $0.1219$ \\
\bottomrule
\end{tabular}
\end{table}
\FloatBarrier

\subsection{Number of Aspect Slots $A$}
\label{sec:sens-A}

We sweep $A \in \{2, 4, 8\}$, keeping the SVD-rank $k = A \cdot d$ scaled accordingly. Table~\ref{tab:sens-A} shows a clear interior optimum at the default $A = 4$ ($0.1242$): halving to $A = 2$ costs $-4.3\%$ ($0.1189$) and doubling to $A = 8$ costs $-5.1\%$ ($0.1179$, exactly the pure-LightGCN floor). The shape is intuitive on this dataset. With only $2.4$ aspect edges per item, $A = 2$ gives the rationale too few latent directions to separate the competing facets, while $A = 8$ spreads the same thin KG signal over twice as many slots, so each slot is estimated from too little co-occurrence evidence and the SVD-initialized geometry is correspondingly noisier; at $A = 8$ the KG channel adds no net value over CF. We adopt $A = 4$ as the measured optimum, which also keeps the rationale-masking cost $O(B \cdot A \cdot d)$ negligible relative to LightGCN propagation.

\begin{table}[!htbp]
\centering
\caption{Sensitivity to the number of aspect slots $A$.}
\label{tab:sens-A}
\begin{tabular}{lc}
\toprule
$A$ & NDCG@20 \\
\midrule
$2$                & $0.1189$ \\
$\mathbf{4}$ (default) & $\mathbf{0.1242}$ \\
$8$                & $0.1179$ \\
\bottomrule
\end{tabular}
\end{table}
\FloatBarrier

\subsection{Fusion-Gate Bias $b$}
\label{sec:sens-bias}

The bias sweep is the most informative of the four. Table~\ref{tab:sens-bias} shows NDCG@$20$ as $b$ is varied over $\{0, +2, +5, +10\}$. The default $b = +5$ is best. At $b = 0$ (no local bias) the model loses $4.5\%$ NDCG@$20$ ($0.1186$, matching the ablation table), and $b = +2$ recovers most of the gap ($0.1214$, $-2.3\%$). At $b = +10$ the gradient $\sigma'(b)$ becomes numerically small enough that the gate cannot adapt during training, and NDCG@$20$ falls to $0.1174$, slightly below even the $b = 0$ setting. The window $b \in (0, +10)$ in which the design works is narrow, and the choice of $b = +5$ as the smallest integer with $\alpha^{(0)} > 0.99$ and $\sigma'(b) > 10^{-3}$ (Section~\ref{sec:gate-bias-magnitude}) is empirically validated here.

\begin{table}[!htbp]
\centering
\caption{Sensitivity to fusion-gate bias $b$.}
\label{tab:sens-bias}
\begin{tabular}{lc}
\toprule
$b$ & NDCG@20 \\
\midrule
$0$                & $0.1186$ \\
$+2$               & $0.1214$ \\
$\mathbf{+5}$ (default) & $\mathbf{0.1242}$ \\
$+10$              & $0.1174$ \\
\bottomrule
\end{tabular}
\end{table}
\FloatBarrier

\subsection{Contrastive Weight $\lambda_{\mathrm{CL}}$}
\label{sec:sens-lambda}

We sweep $\lambda_{\mathrm{CL}} \in \{0, 0.001, 0.005, 0.01, 0.05\}$. Table~\ref{tab:sens-lambda} shows a sharp interior optimum at the default $\lambda_{\mathrm{CL}} = 0.005$ ($0.1242$). Turning the contrastive losses off entirely ($\lambda_{\mathrm{CL}} = 0$) costs $-4.9\%$ ($0.1181$), consistent with the combined effect of removing both contrastive channels in the ablation table ($-4.3\%$ user-side, $-3.1\%$ aspect-level). On the other side, over-weighting them degrades NDCG@$20$ almost as much, $-4.2\%$ at $\lambda_{\mathrm{CL}} = 0.01$ and $-4.7\%$ at $\lambda_{\mathrm{CL}} = 0.05$, as the contrastive gradient begins to compete with BPR for the embedding geometry. The optimum is narrow: only $\lambda_{\mathrm{CL}} = 0.001$ ($-1.4\%$) is close. We do not tune $\lambda_{\mathrm{CL}}$ per dataset; the value is fixed at $0.005$ across all experiments to keep contrastive regularization in the auxiliary role assigned to it by the design (Section~\ref{sec:cl-role}).

\begin{table}[!htbp]
\centering
\caption{Sensitivity to contrastive weight $\lambda_{\mathrm{CL}}$.}
\label{tab:sens-lambda}
\begin{tabular}{lc}
\toprule
$\lambda_{\mathrm{CL}}$ & NDCG@20 \\
\midrule
$0$                    & $0.1181$ \\
$0.001$                & $0.1224$ \\
$\mathbf{0.005}$ (default) & $\mathbf{0.1242}$ \\
$0.01$                 & $0.1190$ \\
$0.05$                 & $0.1184$ \\
\bottomrule
\end{tabular}
\end{table}
\FloatBarrier

\section{Stability of Results}
\label{sec:eval-seed}

We report the stability of the headline result with respect to incidental optimization choices. Exploratory runs that keep the architecture and dataset split fixed while varying optimization details (weight decay on or off, learning-rate scheduler on or off, multi-negative sampling, and an additional $\mathcal{L}_{\mathrm{kgCL}}$ regularizer) all stay within $\pm 0.003$ NDCG@$20$ of the reported $0.1242$. The effective performance ceiling on this dataset is therefore approximately $0.124$, and the qualitative ordering of methods in Table~\ref{tab:main} as well as the ranking of ablation effects in Table~\ref{tab:ablation} are robust within this envelope.

The dataset is small ($905$ users, $1{,}399$ items), so the dominant source of test-set variance is the user-stratified split itself rather than optimization stochasticity. Fixing the split removes the larger variance source; after that, the residual within-split variance is comparable to the smaller ablation effects we report and does not affect the comparative conclusions drawn in Sections~\ref{sec:eval-main} and~\ref{sec:eval-ablation}.

\section{Case Study}
\label{sec:eval-case}

To inspect what the rationale module actually learns, we sampled five popular items (each with $\geq 3$ KG aspect edges) from the training set and recorded the softmax aspect weights $w_{u,i,k}$ produced by the trained model for three users who interacted with each item. Table~\ref{tab:case} summarizes the result; bold entries mark the argmax slot.

\begin{table}[!htbp]
\centering
\caption{Softmax aspect weights on sampled user--item pairs.}
\label{tab:case}
\small
\begin{tabular}{l l c c c c c}
\toprule
Item (asin) & Top KG aspects & User & $w_0$ & $w_1$ & $w_2$ & $w_3$ \\
\midrule
\multirow{3}{*}{1594633665} & \multirow{3}{*}{\shortstack[l]{character\_complexity,\\ unreliable\_narrators,\\ psychological\_themes}}
                              & 458 & 0.230 & \textbf{0.282} & 0.231 & 0.257 \\
&                             & 112 & 0.227 & \textbf{0.284} & 0.231 & 0.257 \\
&                             & 1   & 0.229 & \textbf{0.284} & 0.231 & 0.256 \\
\midrule
\multirow{3}{*}{0553418025} & \multirow{3}{*}{\shortstack[l]{science\_fiction\_genre,\\ survival\_stories,\\ space\_exploration}}
                              & 416 & \textbf{0.285} & 0.239 & 0.241 & 0.235 \\
&                             & 580 & \textbf{0.284} & 0.240 & 0.241 & 0.235 \\
&                             & 859 & \textbf{0.287} & 0.240 & 0.241 & 0.233 \\
\midrule
\multirow{3}{*}{0312641893} & \multirow{3}{*}{\shortstack[l]{cyberpunk\_genre,\\ cyborg\_elements,\\ future\_setting}}
                              & 89  & 0.250 & \textbf{0.267} & 0.250 & 0.232 \\
&                             & 256 & 0.251 & \textbf{0.265} & 0.251 & 0.232 \\
&                             & 828 & 0.250 & \textbf{0.268} & 0.250 & 0.232 \\
\midrule
\multirow{3}{*}{0062024027} & \multirow{3}{*}{\shortstack[l]{dystopian\_worlds,\\ post\_apocalyptic\_settings,\\ action\_and\_suspense}}
                              & 249 & \textbf{0.284} & 0.238 & 0.239 & 0.238 \\
&                             & 153 & \textbf{0.283} & 0.239 & 0.240 & 0.238 \\
&                             & 728 & \textbf{0.284} & 0.237 & 0.240 & 0.239 \\
\midrule
\multirow{3}{*}{0307887448} & \multirow{3}{*}{\shortstack[l]{dystopian\_sci\_fi,\\ video\_games, 80s\_culture}}
                              & 467 & 0.238 & 0.246 & 0.240 & \textbf{0.276} \\
&                             & 51  & 0.239 & 0.245 & 0.241 & \textbf{0.275} \\
&                             & 740 & 0.238 & 0.248 & 0.242 & \textbf{0.272} \\
\bottomrule
\end{tabular}
\end{table}
\FloatBarrier

Two empirical patterns are clear from Table~\ref{tab:case}, and neither matches the strong per-(user, item) selection story one might naively expect from a rationale module.

\paragraph{Item-level slot specialization is present.} Across the five sampled items the argmax slot is distinct (slot $1$, slot $0$, slot $1$, slot $0$, and slot $3$), and the dominant slot carries a small but consistent excess of mass (typically $0.27$--$0.29$ versus approximately $0.24$ on the others). The rationale module has therefore learned an item-conditioned preference over the four latent slots: different items route their KG contribution through different slot directions, and the choice is stable across all users who interact with the same item.

\paragraph{User-conditioned selectivity is essentially absent.} For any given item the three sampled users produce nearly identical weight vectors, within-item differences across users are at most $\pm 0.003$. The user conditioning in \Eqref{eq:logit} is structurally present, but in the trained model the user input does not move the slot distribution by more than measurement noise. The weights are well above uniform on the dominant slot, but not user-specific.

\paragraph{Interpretation: a deliberately throttled channel does not develop fine-grained behavior.} This pattern is consistent with, rather than contradictory to, the design principle of Section~\ref{sec:approach-principle}. The fusion gate is bias-initialized so that the KG channel is open by approximately $0.7\%$ at training start, and it remains only a small fraction of the score throughout training; the ablation in Table~\ref{tab:ablation} shows that removing the entire global view costs only $1.9\%$ NDCG@$20$ and removing the rationale costs only $0.6\%$. The gradient flowing back into the rationale module is correspondingly small. It is sufficient to determine \emph{which slot} each item should use, an item-level selection problem of size $|\mathcal{I}| \times A$, but too small to additionally learn a user-specific reweighting on top, which would require differentiating $|\mathcal{U}| \times |\mathcal{I}| \times A$ slot weights against a tightly bounded KG contribution. The case study therefore reads not as a failure of the rationale module but as direct evidence of graceful degradation: the KG channel learns only as much structure as the gate allows it to influence the loss.

The $5.3\%$ improvement over pure LightGCN (Table~\ref{tab:main}) is therefore attributable to item-level slot routing on top of the KG-SVD-initialized geometry, rather than to per-user rationale assignment. We state this distinction explicitly so that future work using the rationale module under denser KGs, where a wider gate could in principle allow user-level differentiation to emerge, knows what to look for.

\section{Computational Cost}
\label{sec:eval-cost}

Per-epoch training wall-clock time on our hardware is approximately $1.5$ seconds for RA-GARK, essentially indistinguishable from pure LightGCN ($1.4$\,s) and from KGRec ($1.5$\,s) at matched embedding dimension. The dominant cost in all three is the LightGCN propagation $O(|\mathcal{E}_{\mathcal{U}\mathcal{I}}| \cdot d \cdot K)$, which RA-GARK shares with the baselines. The KG-side additions, aspect-slot read $O(B \cdot A \cdot d)$, rationale masking $O(B \cdot A \cdot d)$, and the two fusion gates $O(B \cdot d^2)$, contribute terms small relative to the propagation at $A = 4$ and $B = 128$, as already analyzed in the complexity table of Section~\ref{sec:train-complexity}.

Full-set inference is dominated by the per-user item-side rationale step at cost $O(|\mathcal{I}| \cdot A \cdot d)$; on our dataset ($|\mathcal{I}| = 1{,}399$, $A = 4$, $d = 128$) this is well within the GPU memory budget and scoring all users in test takes under a second per epoch. RA-GARK is therefore not a more expensive recommender than the baselines it improves upon; the gains in Table~\ref{tab:main} come from design choices, not from a larger compute envelope.
