\chapter{Conclusion and Future Work}
\label{ch:conclusion}

\section{Conclusion}
\label{sec:conclude-main}

This thesis began from a specific empirical anomaly: on the sparse review-derived knowledge graph used here, four representative KG-aware recommenders (KGAT, KGCL, MCCLK, KGRec) are uniformly outperformed by pure LightGCN (Chapter~\ref{ch:intro}, reproduced in Section~\ref{sec:eval-main}). The architectural cause shared by these baselines, we argued, is that the KG is wired into the scoring pipeline as a mandatory component: KG embeddings participate in message passing, and there is no structural switch by which the model can disengage the KG when it turns out to be unreliable. The same noise that one hopes to harvest as semantic signal therefore enters the gradient that produces the collaborative representation, and below some KG-quality threshold this trade becomes net negative.

Our response, RA-GARK (Chapter~\ref{ch:methodology}), is organized around a single design principle (Section~\ref{sec:approach-principle}): the KG should be a \emph{side channel} whose contribution to the final score is determined by a learned gate, biased at initialization toward a safe default (pure collaborative filtering) and opened only when training signal indicates that the KG channel reduces loss. Three concrete design choices implement this principle and were measured individually in Section~\ref{sec:eval-ablation}:

\begin{itemize}
  \item \textbf{KG-SVD initialization} of the per-item aspect slots, which preserves the angular geometry of the aspect-induced semantic subspace at the start of training. Removing it costs $5.3\%$ NDCG@$20$.
  \item \textbf{Softmax rationale masking} over the $A$ latent slots, which encodes the assumption that a user's preference for an item is dominated by a small number of competing semantic facets. Replacing the cross-aspect softmax with an element-wise sigmoid costs $7.0\%$ NDCG@$20$, the single largest individual ablation in this thesis.
  \item \textbf{Local-biased fusion gate}, whose scalar bias $b = +5$ is initialized so that $\alpha^{(0)} \approx 0.993$ and RA-GARK begins training as essentially pure LightGCN. Reverting to $b = 0$ costs $5.3\%$ NDCG@$20$.
\end{itemize}

With all three in place, RA-GARK achieves NDCG@$20 = 0.1238$ on our benchmark, an improvement of $+13.1\%$ over the strongest KG-aware baseline (KGRec) and $+5.0\%$ over pure LightGCN, at essentially the same per-epoch wall-clock cost (Section~\ref{sec:eval-cost}). The $+5.0\%$ figure is the methodologically more interesting of the two: it shows that on the same data on which four published KG-aware methods turn the KG signal into a net liability, RA-GARK's design choices turn it back into a net asset.

The single-sentence summary of the thesis is therefore: \emph{when the KG is unreliable, what an architecture most needs is not a better KG aggregator but a structural switch by which the KG can be opted out of.}

The value of the thesis lies less in the specific numbers than in this reframing. It locates the difficulty of sparse-KG recommendation in the architecture rather than in the KG itself: the same KG that is a liability under mandatory fusion becomes an asset once its use is made structurally optional, with no change to the KG. Read this way, graceful degradation, biasing an auxiliary channel toward a safe default and opening it only when training signal warrants, is a reusable principle for integrating any signal whose reliability cannot be assumed in advance, rather than a device specific to review-derived knowledge graphs. Whether that principle holds beyond the single sparse benchmark studied here is the empirical question taken up in Section~\ref{sec:conclude-future}.

\section{Methodological Insight: Attention Normalization}
\label{sec:conclude-insight}

The largest single-component effect we observe in this thesis is not in any of the new architectural pieces RA-GARK contributes, but in a choice that is treated as a tuning detail in much of the recommendation literature: whether the attention head in the rationale module uses softmax (cross-aspect competition) or sigmoid (per-aspect independence). On our benchmark, replacing softmax with sigmoid in this single module costs $7.0\%$ NDCG@$20$ (Section~\ref{sec:ablation-softmax}), which is large relative to the $1$--$3\%$ swings typically produced by ordinary hyperparameter tuning.

The recommendation literature offers no consensus on this choice. DIN~\cite{zhou2018din} uses per-target sigmoid attention over historical items; NAIS~\cite{he2018nais} and AFM~\cite{xiao2017afm} use softmax-normalized attention over feature interactions; KGRec~\cite{yang2023kgrec} uses an edge-level Bernoulli decision that is structurally closer to sigmoid. Each of these choices is internally consistent with the semantics of its target task. Our case-study inspection (Section~\ref{sec:eval-case}) showed that, in our setting, the trained softmax weights are not near-one-hot but only softly peaked ($\max_k w_k \approx 0.28$ over four slots), so the operative mechanism cannot be one-hot competitive selection. The property that softmax provides and sigmoid does not is rather a structural one: the sum-to-one normalization keeps the global-view item embedding's magnitude bounded by $\max_k \|\mathbf{a}_{i,k}\|$ regardless of logit scale, whereas sigmoid produces a magnitude that drifts with the logits. In a design where the KG channel is deliberately throttled by the fusion gate (Section~\ref{sec:gate-bias}), keeping the channel's contribution bounded and predictable is precisely what the gate's calibration depends on, and the magnitude drift introduced by sigmoid is what we believe causes the $7.0\%$ degradation.

We frame this as a hypothesis rather than a methodological claim. The evidence comes from a single dataset, and the effect's size and mechanism on other datasets and on differently-throttled architectures are unknown. The hypothesis is concrete enough to be falsified by replication: \emph{when a rationale-like attention head feeds an auxiliary channel whose contribution is held to a small fraction of the score by a fusion gate, the load-bearing property is the head's magnitude control (sum-to-one normalization) rather than one-hot competition; replacing the softmax head with an independent sigmoid head will then degrade ranking quality by a margin substantially larger than ordinary hyperparameter noise.} If this holds beyond our setting, normalization-vs-competition becomes an explicit design axis in rationale-aware architectures, not a tuning detail.

\section{Limitations}
\label{sec:conclude-limits}

The contributions of this thesis are bounded by three limitations, which we state explicitly so that the scope of the claims is unambiguous.

\paragraph{Single sparse benchmark.} All quantitative results in this thesis are reported on one dataset: a subset of Amazon Books reviews with $905$ users, $1{,}399$ items, and an item--aspect KG containing $3{,}370$ edges over $2{,}098$ aspect tags (Section~\ref{sec:eval-dataset}). The benchmark was chosen because it exhibits the empirical anomaly that motivates the thesis, but no claim is made that RA-GARK's design choices transfer unchanged to other recommendation datasets or other KG schemas. In particular, the headline observation, that pure LightGCN beats all four KG-aware baselines we tested, need not hold on datasets where the KG is denser, more curated, or more aligned with the recommendation signal.

\paragraph{KG construction is prior work.} The item--aspect KG itself is built by the pipeline of~\cite{ho2024reviewkg}, which combines visual and textual encoders (ResNet, Qwen-VL, BART, Mistral) to extract review-derived aspect mentions. We use this KG as given and do not propose any modification to the construction process. The contributions of this thesis lie entirely on the modeling side, taking the KG as fixed input.

\paragraph{Dense-KG regime not validated.} RA-GARK was designed for, and evaluated on, the sparse-KG setting in which the KG signal is unreliable. We make no claim about its behavior when the KG is dense and accurate. In that setting, the traditional home turf of KGAT and KGIN, a bias-initialized gate that defaults to ``KG off'' is precisely the wrong inductive prior, and the deep-fusion baselines we critique here may well be the appropriate choice. Section~\ref{sec:conclude-future} discusses how the gate would need to be reconfigured to handle a symmetric regime.

\section{Future Work}
\label{sec:conclude-future}

Four directions follow naturally from the contributions and limitations above.

\paragraph{Multi-dataset replication of the attention-normalization finding.} The methodological hypothesis articulated in Section~\ref{sec:conclude-insight} is concrete enough to test on other rationale-aware recommendation benchmarks, on other item-attribute schemas, and on rationale modules in adjacent domains (sequential recommendation, conversational recommendation). A clean replication, holding the dataset and the rest of the model fixed and toggling only the rationale head's normalization, would either elevate this from a single-dataset observation to a general design guideline, or sharpen the conditions under which it holds.

\paragraph{Dense-KG validation.} As noted under limitations, RA-GARK has not been evaluated on datasets where the KG is dense and well-curated. The natural follow-up is to run RA-GARK alongside the KG-aware baselines on standard dense-KG benchmarks (e.g.\ Amazon-Book / Last-FM with Freebase-derived KGs). The expected outcome is that the local-biased fusion gate gracefully degrades to behavior comparable to pure LightGCN with a small KG contribution, rather than collapsing; whether RA-GARK matches deep-fusion baselines in that setting is an empirical question worth answering.

\paragraph{Cross-domain transfer of the gating principle.} The design idea behind the local-biased fusion gate, biasing the safe default at initialization and letting training open the auxiliary channel only when its contribution reduces loss, is not specific to KG-aware recommendation. The same construction transfers, in principle, to any setting in which a primary signal is robust and an auxiliary signal is high-variance: cold-start with side information, multi-modal fusion when one modality is noisy, ensemble-style late fusion of heterogeneous predictors. Whether the headline effect we observe survives such transfers is open.

\paragraph{Reverse-sparse regime: sparse CF with dense KG.} The symmetric inverse of the setting studied in this thesis is the situation in which the user--item interaction graph is sparse (e.g.\ cold-start, new user, new domain) while the KG is dense and reliable. There, the safe default that the gate should encode is the opposite of the one we use here: LightGCN is now the unreliable side, and the KG-derived global view should be the channel that is open by default. This is the traditional setting of KGAT, KGIN, and the dense-KG line of work; our analysis suggests that the principle of \emph{biasing the safe default at initialization} is the design move that transfers across the two regimes, while the specific direction of the bias is dataset-determined.

Concretely, redesigning RA-GARK for this setting requires three coordinated changes: the fusion bias should flip from $+5$ to a negative value so the KG channel is open by default; the KG-SVD step's role may shift from initialization to a primary inference path; and the local view likely needs a different regularizer to remain useful when interaction signal is scarce. A full architectural treatment of the reverse regime requires redesigning both the local view and the gate direction together, and is left to future work.
